A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer depending on primary tumor location

### 2024-10-03 _tex/manuscript.tex
Built site for gh-pages

\begin{equation}\phantomsection\label{eq-likelihood-HMM}{
P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_i \left[P \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}, T_i \right) \right]^{N_{i}}
}\end{equation}

In equation~\ref{eq-likelihood-HMM} we take the likelihood of observing
a patient in t-stage \(T_i\) and with involvement \(\xi_i\) to the power
of \(N_i\), corresponding to the number of observations of the specific
diagnosis \(i\) in the dataset \(\mathbf{D}\). By multiplying over all
diagnoses \(i\) we get the likelihood of observing dataset
\(\mathbf{D}\) given the previously intoduced model with parameters
\(\boldsymbol{\theta}\)

@@

Assume we have a dataset \(\mathbf{D}\) comprising the number of
patients \(N_{is}\) diagnosed with LNL involvement state \(i\) and
having a primary tutor in subsite \(s\). Let us further assume that we
want to describe this dataset using a mixture of \(M\) HMMs, each with a
different set of model parameters \(\boldsymbol{\theta_m}\).

As the generative model of the data, we assume that a patient with
subsite \(s\) is generated with probability \(\pi_m^s\) from model
\(m\),which satisfies the constraints \(\sum_{m = 1}^{M}\pi_m^s = 1\)
and \(\pi_m^s \geq 0\) for all \(m\) and all subsites \(s\).

The likelihood of the dataset is expressed as:

\begin{equation}\phantomsection\label{eq-mixture-distribution}{
P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_s \prod_i \left[ \sum_{m=1}^M \pi_m^s P_m \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}_m, T_i \right) \right]^{N_{is}}
}\end{equation}

In equation~\ref{eq-mixture-distribution} we compute the probability of
each state \(\boldsymbol{\xi}_i\) under model \(m\) similarly to
equation~\ref{eq-likelihood-HMM}. However, we weight each model with the
mixing parameter \(\pi_m^s\). Since the mixing parameters are subsite
dependent, we differ between each subsite.

@@

In the E-step, we compute the expected value of the latent variables
\(\boldsymbol{\epsilon}\), which represent the probability that each
observation \(\mathbf{X} = \boldsymbol{\xi}_i\) originates from model
\(m\). Given the current estimates of \(\boldsymbol{\theta}\) and
\(\boldsymbol{\pi}\), the expectation value \(\gamma(\epsilon_m^i)\) is
calculated as:

\begin{equation}\phantomsection\label{eq-expectation-step}{
\mathbb{E}[\epsilon_m] = \gamma(\epsilon_m^i) = \frac{\pi_m^s P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_m}, T_i)}{\sum_{j=1}^M \pi_j^s P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_j}, T_i)}
}\end{equation}

Here, the mixing parameter \(\pi_m^s\) depends on the subsite of
diagnosis, \(Z_i\), as denoted by the superscript \(s\).

@@

In the M-step, the model parameters are updated by maximizing the
following likelihood, based on the expectations computed in the E-step:

\begin{equation}\phantomsection\label{eq-maximization-step}{
\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}, \boldsymbol{\pi}) = \sum_{i=1}^{N} \sum_{m=1}^M \gamma(\epsilon_m^i) \left( \ln \pi_m^s + \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_m}, T_i) \right)
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-mixture-maximization}{
\pi_m^s = \frac{1}{N^s} \sum_i^{N^s} \gamma(\epsilon_m^i)
}\end{equation}

@@

\[
\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}_m) = \sum_{i=1}^{N} \gamma(\epsilon_m^i) \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \theta_m, T_i)
\]

By iterating these steps, the EM algorithm converges to a (local)
maximum of the likelihood.

\begin{equation}\phantomsection\label{eq-mixture-llh}{
\ln p(\mathbf{D}|\boldsymbol{\theta},\boldsymbol{\pi}) = \left\{\sum_{\boldsymbol{\epsilon}} p(\mathbf{D},\boldsymbol{\epsilon}|\boldsymbol{\theta}, \boldsymbol{\pi})\right\}
}\end{equation}

>>>>

\begin{equation}\phantomsection\label{eq-likelihood-HMM}{
P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_i \prod_j \left[P \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}, T_j \right) \right]^{N_{ij}}
}\end{equation}

In equation~\ref{eq-likelihood-HMM} we take the likelihood of observing
a patient in t-stage \(T_j\) and with involvement \(\xi_i\) to the power
of \(N_{ij}\), corresponding to the number of observations of the
patients in state \(i\) and t-stage \(T_j\) (either early or late) in
the dataset \(\mathbf{D}\).

@@

The mixture model assumes that the data is generated by a set of \(M\)
different underlying models. In our case, we introduce \(M\) distinct
lymphatic spread models. Each patient \(k\) is assumed to be generated
by one specific model from this set of \(M\) models.

To formalize this, each patient \(k\) can be assigned to a specific
model via a binary latent vector \(\boldsymbol{\epsilon}_k\). The vector
\(\boldsymbol{\epsilon}_k\) has length \(M\), with exactly one element
set to 1, indicating which model generated the patient's data, and all
other elements set to 0. Thus, for patient \(k\), the latent variable
\(\boldsymbol{\epsilon}_k\) can be interpreted as a categorical
indicator variable that encodes the assignment to one of the \(M\)
lymphatic spread models.

Next, we introduce subsite-specific mixing parameters
\(\boldsymbol{\pi}^s = \{\pi_1^s, \pi_2^s, \dots, \pi_M^s\}\), where
each \(\pi_m^s\) represents the prior probability that a patient with
tumors in subsite \(s\) is generated by model \(m\). These mixing
parameters must satisfy the condition:

\[
\sum_{m=1}^M \pi_m^s = 1, \quad \forall s
\]

The mixing parameters \(\boldsymbol{\pi}^s\) determine the proportion of
patients with tumors in subsite \(s\) that are expected to be generated
by each model. Specifically, \(\pi_m^s\) is the probability that a
randomly chosen patient with a tumor in subsite \(s\) is generated by
model \(m\).

The latent variables \(\boldsymbol{\epsilon}_{k}\), on the other hand,
are patient-specific and represent the assignment of each patient to a
particular model. We will later generalize patients with the same
diagnosis to the latent variable \(\boldsymbol{\epsilon}_{i,j}\). The
subscripts \(i\) and \(j\) represent that the latent variables are only
dependent on the diagnosis of the patient, which consists of lymphatic
involvement \(\boldsymbol{\xi}_i\) and t-stage \(T_j\). The latent
variables are central to the model, as they are hidden (unobserved) and
must be inferred.

The joint probability of the observed data \(\mathbf{D}\) (i.e., patient
data) and the latent variables \(\boldsymbol{\epsilon}\) is given by:

\begin{equation}\phantomsection\label{eq-complete-likelihood}{
P(\mathbf{D}, \boldsymbol{\epsilon} | \boldsymbol{\theta}, \boldsymbol{\pi}) = \prod_{k=1}^{K} \prod_{m=1}^{M} \left[ \pi_m^s P(\mathbf{D}_k | \boldsymbol{\theta}_m, \boldsymbol{\pi}) \right]^{\epsilon_{k}^m}
}\end{equation}

Here:

\begin{itemize}
\tightlist
\item
  \(\pi_m^s\) is the mixing parameter for subsite \(s\) and model \(m\),
\item
  \(P(\mathbf{D}_k | \boldsymbol{\theta}_m)\) is the likelihood of
  patient \(k\)'s data given that it was generated by model \(m\),
\item
  \(\boldsymbol{\theta}_m\) represents the parameters of model \(m\),
\item
  \(\epsilon_{k}^m\) is the latent variable that indicates whether
  patient \(k\) was generated by model \(m\).
\end{itemize}

\textbf{Incomplete data likelihood which can probably be skipped}

Since the latent variables \(\boldsymbol{\epsilon}_k\) are unobserved,
the incomplete complete data likelihood marginalizes over all possible
latent assignments to reflect the uncertainty about which model
generated each patient's data:

\[
P(\mathbf{D} | \boldsymbol{\theta}, \boldsymbol{\pi}) = \sum_{\boldsymbol{\epsilon}} P(\mathbf{D}, \boldsymbol{\epsilon} | \boldsymbol{\theta}, \boldsymbol{\pi})
\]

This allows us to integrate out the latent variables
\(\boldsymbol{\epsilon}\), as their true values are unknown.

\textbf{Continues from here}

The latent variables \(\boldsymbol{\epsilon}_k\) are unknown and must be
inferred from the observed data. The mixture model assumes that each
patient is generated by exactly one of the \(M\) models. However, in our
application, this assumption does not reflect continuous or overlapping
behaviors between subsites. Instead, we infer a probabilistic
distribution over the latent variables, given the observed data and
model parameters. The posterior distribution:

\[
P(\boldsymbol{\epsilon}_k | \mathbf{D}_k, \boldsymbol{\theta}, \boldsymbol{\pi})
\]

describes the probability that patient \(k\) was generated by each of
the \(M\) models, allowing for soft assignments where each patient is
associated with multiple models with certain probabilities, as opposed
to being strictly assigned to just one. Given the probability
distribution for the latent variables \(\boldsymbol{\epsilon}_k\), we
compute the expectation value
\(\mathbb{E}[\boldsymbol{\epsilon}_k] = \boldsymbol{\gamma}(\boldsymbol{\epsilon}_k)\)
for each patient to then apply the complete data likelihood given in
equation~\ref{eq-complete-likelihood}.

We can now apply the mixture model to our dataset \(\mathbf{D}\)
comprising the number of patients \(N_{ijs}\) diagnosed with LNL
involvement state \(\mathbf{X} = \boldsymbol{\xi}_i\), t-stage \(T_j\),
with a primary tumor in subsite \(s\).

The likelihood of the complete dataset is then expressed as:

\begin{equation}\phantomsection\label{eq-complete-likelihood-explicite}{
P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_i \prod_j \prod_s \left[ \prod_{m=1}^M \left[ \pi_m^s P_m \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}_m, T_j \right) \right]^{\gamma(\epsilon_{i,j}^m)} \right]^{N_{ijs}}
}\end{equation}

In equation~\ref{eq-complete-likelihood} we compute the likelihood of
observing state \(\boldsymbol{\xi}_i\) with t-stage \(T_j\) under model
\(m\) similarly to equation~\ref{eq-likelihood-HMM}. However, we weight
each model with the mixing parameter \(\pi_m^s\). Since the mixing
parameters are subsite dependent, we differ between each subsite.

@@

In the E-step, we compute the expected value of the latent variables
\(\boldsymbol{\epsilon}\), which represent the probabilities of a
patient \(k\) originating from one of the models \(m\). Given the
current estimates of \(\boldsymbol{\theta}\) and \(\boldsymbol{\pi}\),
the expectation value \(\gamma(\epsilon_m^k)\) is calculated as:

\begin{equation}\phantomsection\label{eq-expectation-step}{
\mathbb{E}[\epsilon^m] = \gamma(\epsilon^m_{i,j}) = \frac{\pi_m^s P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_m}, T_j)}{\sum_{n=1}^M \pi_n^s P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_n}, T_j)}
}\end{equation}

THE OTHER OPTION

\begin{equation}\phantomsection\label{eq-expectation-step2}{
\mathbb{E}[\epsilon^m] = \gamma(\epsilon^m_k) = \frac{\pi_m^s P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_m}, T_k)}{\sum_{j=1}^M \pi_j^s P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_j}, T_k)}
}\end{equation}

Here, the mixing parameter \(\pi_m^s\) depends on the subsite location
of the primary tumor of patient \(k\).

@@

In the M-step, the model parameters are updated by maximizing the
complete data likelihood in equation
equation~\ref{eq-complete-likelihood-explicite}. For better handling, we
are optimizing the log likelihood:

\begin{equation}\phantomsection\label{eq-maximization-step}{
\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}, \boldsymbol{\pi}) = \sum_{i} \sum_j \sum_{s} N_{i,j,s} \sum_{m=1}^M \gamma(\epsilon^m_{i,j}) \left( \ln \pi_m^s + \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_m}, T_j) \right)
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-mixture-maximization}{
\pi_m^s = \frac{1}{N^s} \sum_k^{N^s} \gamma(\epsilon^m_k)
}\end{equation}

@@

\[
\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}_m) = \sum_{i} \sum_j \gamma(\epsilon^m_{i,j}) \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \theta_m, T_j)
\]

By iterating these steps, the EM algorithm converges to a (local)
maximum of the complete data likelihood
equation~\ref{eq-complete-likelihood}.
